\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-09-15 challenge.tex
intro
Challenge#1

\subsection{Challenges}

>>>>

Supporting inheritance and family polymorphism in a proof assistant poses challenges
far beyond those encountered in an object-oriented setting~\cite{ncm2004,vc-calculus-2006,zm2017},
because the underlying programming language is simultaneously
functional, dependently typed, a logic, and an interactive tool.

\noindentparagraph{Extensible inductive types vs.\ exhaustive induction reasoning.}

Inductive types, generalizing algebraic data types found in
functional languages, are a central feature of any proof assistant being
used for mechanizing language metatheories.
They offer a means to define abstract syntax, typing rules, reduction
rules, and so on.
Unfortunately, inductive types are closed to extension.

A family-polymorphism mechanism could potentially support extensible
inductive types, by allowing a \emph{derived family} to add new constructors
to inductive types inherited from a \emph{base family}.
Such capability would be useful for extending mechanized languages.
For example, \cref{fig:?} shows the part of an STLC extension that could
potentially be mechanized using extensible inductive types: a derived family
inherits constructors of inductive types from the base family (viz.\ the $\cdots$'s)
and adds new constructors to model iso-recursive types.

\begin{figure}
\fontsize{8.5}{9}\selectfont
\def\MathparLineskip{\lineskip=0.50ex}

\fcolorbox{black}{black!10}{
\begin{minipage}{0.38\textwidth}

\noindent%
Types \quad
$ \tau \ \Coloneqq\ \mathbb{1} \bnf  \tau_1 \to \tau_2 $
\medskip

Terms \quad
$ e \ \Coloneqq\ x \bnf () \bnf \lambda x.\,e \bnf e_1~e_2 $
\medskip

Typing rules
\begin{mathpar}
\Rule{ \Gamma(x)=\tau }{ \Gamma \vdash x : \tau }

\Rule{ \Gamma, x:\tau_1 \vdash e : \tau_2 }{ \Gamma \vdash \lambda x.\,e : \tau_1 \to \tau_2 }

\Rule{}{ \Gamma \vdash () : \mathbb{1} }
\quad
\Rule{
  \Gamma \vdash e_1 : \tau_1 \to \tau_2
  \quad
  \Gamma \vdash e_2 : \tau_1
}{
  \Gamma \vdash e_1~e_2 : \tau_2
}

\end{mathpar}

Values
\begin{mathpar}
\Rule{}{\textit{Val}(())}

\Rule{}{\textit{Val}(\lambda x.\,e)}
\end{mathpar}

Reduction rules
\begin{mathpar}
\Rule{\textit{Val}(e_2)}{ (\lambda x.\,e_1)~e_2 \longrightarrow \subst{e_2}{e_1}{x} }

\Rule{ e_1 \longrightarrow e_1' }{ e_1~e_2 \longrightarrow e_1'~e_2 }

\Rule{ \textit{Val}(e_1) \quad e_2 \longrightarrow e_2' }{ e_1~e_2 \longrightarrow e_1~e_2' }

\end{mathpar}

\end{minipage}
}%
%\tikz \draw [arrows = {-Latex[width'=0pt .5, length=10pt, fill=white]}] (0,0) -- (1,0);
\tikz \draw [
  line width = .5pt,
  arrows = {-Stealth[inset=0pt, length=5pt, fill=white, angle'=60, scale=2]}
] (1,0) -- (0,0);%
%https://tikz.dev/tikz-arrows
\fcolorbox{black}{black!10}{
\begin{minipage}{0.49\textwidth}

\noindent%
Types \quad $ \tau \ \Coloneqq\ \cdots \bnf \alpha \bnf \mu\alpha.\,\tau$
\medskip

Terms \quad $ e \ \Coloneqq\ \cdots \bnf \mathrm{fold}\;e \bnf \mathrm{unfold}\;e $
\medskip

Typing rules
\begin{mathpar}
\cdots
\quad
\Rule{
  \Gamma \vdash e : \subst{\tau}{\mu\alpha.\,\tau}{\alpha}
}{
  \Gamma \vdash \mathrm{fold}\;e : \mu\alpha.\,\tau
}
\quad
\Rule{
  \Gamma \vdash e : \mu\alpha.\,\tau
}{
  \Gamma \vdash \mathrm{unfold}\;e : \subst{\tau}{\mu\alpha.\,\tau}{\alpha}
}

\end{mathpar}

Values
\begin{mathpar}
\cdots

\Rule{ \textit{Val}(e) }{ \textit{Val}(\mathrm{fold}\;e) }

\end{mathpar}

Reduction rules
\begin{mathpar}
\cdots

\Rule{ e \longrightarrow e' }{ \mathrm{unfold}\;e \longrightarrow \mathrm{unfold}\;e' }

\Rule{ e \longrightarrow e' }{ \mathrm{fold}\;e \longrightarrow \mathrm{fold}\;e' }

\Rule{ \textit{Val}(e) }{ \mathrm{unfold}\;(\mathrm{fold}\;e) \longrightarrow e }

\end{mathpar}

\end{minipage}
}

\caption{%
  Statics and dynamics of STLC (left) and its extension with iso-recursive types (right).
}
\label{fig:?}
\end{figure}

However, there is a tension between extensibility of inductive types
and exhaustivity of induction reasoning.
The substitution function and the substitution lemma are examples that
require induction reasoning.
In the derived family, to ensure that induction remains exhaustive,
how could the programmer account for the new cases arising from the new
constructors, without redefining or forcing the rechecking of those cases
already handled by the base family?

The tension results from the duality between variants and records.
Existing family-polymorphism mechanisms are designed for record-like
language constructs like classes and objects; inductive types, by
contrast, are variant-like.

\noindentparagraph{Late binding vs.\ dependent typing.}

\noindentparagraph{Essence of inheritance vs.\ consistency of a logic.}

\noindentparagraph{Family vs.\ interactivity.}

\newpage

### 2022-09-15 coq-example.tex
Challenge#1

And at the end, the compiled module for "LangTyping" and "LangCP" is put into tests over simple queries. These "LangTyping" and "LangCP" also illustrates that we still have
computability in the presence of family.\YZ{Can there be a diagram (a la Figure 1 in Familia paper) illustrating the relationships between (and functionalities of) the different families?}\EDJreply{Done. Please check}

>>>>

And at the end, the compiled module for "LangTyping" and "LangCP" is put into tests over simple queries. These "LangTyping" and "LangCP" also illustrates that we still have
computability in the presence of family.\YZ{Mention the alternative approach of using parameterized modules.}

### 2022-09-15 introduction.tex
Challenge#1

Language mechanisms including
mixins~\cite{mixin-1990},
virtual classes~\cite{virtualclasses-1989,vc-calculus-2006},
virtual types~\cite{thorup97} and associated types~\cite{ckj05},
extensible cases~\cite{bac2006},
and so on, are all based on this essential idea of inheritance.
%
In particular, when a language mechanism enables inheritance over
recursive definitions containing related types and terms,
it is said to support \emph{family polymorphism}~\cite{ernst2001family}:
types and terms are polymorphic to a family they are nested within.

@@

We contribute a language design that integrates family polymorphism into
a proof assistant.
Because code and proofs are polymorphic to a family they are nested
within, they can be

\begin{wrapfigure}[6]{r}{.215\textwidth}
\includegraphics[scale=.48]{graphics/stlc-intro.pdf}
\end{wrapfigure}

\noindent
inherited and reused by a derived family—%
family polymorphism enables extensible metatheory mechanization.

%As an example, consider \cref{fig:STLC-example}, the simply typed
%lambda calculus (STLC) mechanized in Coq.
%A programming challenge there is to define distinct extensions of this
%STLC formalization to support distinct features,
%and then selectively compose these extensions to form new STLC variants.
As an example, the diagram to the right depicts an extensible
mechanization of the simply typed λ-calculus (STLC), using family
polymorphism.
Different extensions to STLC can all inherit from the base STLC family:
they reuse mechanized metatheories, from syntax to the type safety
theorem, only adding new constructors to inductive types
and new cases to the induction proofs and recursive functions as needed
by an extension.

@@

\item We present case studies of using our Coq plugin to mechanize
language metatheories.
They show how our language design naturally solves the EP and
enables a high degree of reuse and extensibility
for proof engineering.

>>>>

Language mechanisms including
mixins~\cite{mixin-1990},
virtual classes~\cite{virtualclasses-1989,vc-calculus-2006},
virtual types~\cite{thorup97} and associated types~\cite{ckj05},
extensible cases~\cite{bac2006},
and so on, are all based on this essential idea of inheritance.
%
In particular, when a language mechanism allows inheritance over
recursive definitions containing related types and terms,
it is said to support \emph{family polymorphism}~\cite{ernst2001family}:
types and terms are polymorphic to a family they are nested within.

We contribute a language design that integrates family polymorphism into
a proof assistant.
Because code and proofs are polymorphic to a family they are nested
within, they can be
inherited and reused by a derived family.
Hence, family polymorphism allows for extensible metatheory mechanization.

@@

\begin{wrapfigure}[6]{r}{.200\textwidth}
\includegraphics[scale=.48]{graphics/stlc-intro.pdf}
\end{wrapfigure}

%\noindent
%As an example, consider \cref{fig:STLC-example}, the simply typed
%lambda calculus (STLC) mechanized in Coq.
%A programming challenge there is to define distinct extensions of this
%STLC formalization to support distinct features,
%and then selectively compose these extensions to form new STLC variants.
As an example, the diagram to the right depicts an extensible
mechanization of the simply typed λ-calculus (STLC), using family
polymorphism.
An extension of STLC with products and another with iso-recursive types
can both inherit from the base STLC family:
they reuse mechanized metatheories,
from abstract syntax all the way to the type-safety theorem,
only adding new constructors to inductive types
and adding new cases to recursive functions and induction proofs
\emph{as needed} by an extension.

@@

\item We present case studies of using our Coq plugin to mechanize
language metatheories (\cref{sec:coqexample}).
They show how our language design naturally solves the EP and
enables a high degree of reuse and extensibility
for proof engineering.
